Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-05 latex/main.tex
Some tex fixups

\subsection*{Quantification of evidence for recombination}
%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.
Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination, as measured by the number of resulting 
mutations (Figure~\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination, the stronger the 
evidence for recombination.
We calculated the number of averted mutations for all the ``robust'' recombination 
events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages}) and those studied by Jackson et
al.\cite{Jackson2021}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{fig_recombinant_evidence.pdf}
    \caption{Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes, colored according
to their classification by rebar (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labelled starting with `X' and
those identified by \cite{Jackson2021} (see Table \ref{tab:jackson}) are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1). 
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
    \label{fig:averted_mutations}
\end{figure}

\subsection*{Recombination breakpoint intervals along the genome}

We explored the spatial distribution of the breakpoint locations of
the ``robust'' recombination events along the genome, and quantified how precisely breakpoints were localized.
For each event, we computed an interval estimate of its breakpoint location,
using its size as a measure of precision (median, 2,207 bases; STAR Method).

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene (\ref{fig:recombinant_intervals}),
consistent with the pattern of excess observed in previous studies \cite{Turakhia2022,Li2024}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
allowing recombination endpoints to be estimated more precisely.
Reassuringly, sites having unusually high mutation counts are absent in this region (\ref{fig:recombinant_intervals}),
which would confound recombination detection.

Furthermore, we find a negative relationship between
breakpoint interval size and the amount of divergence between recombinant parents,
which is measured by time to their MRCA (\ref{fig:recombinant_intervals}).
When the two parental lineages have diverged for longer periods of time (the rightmost points),
more information becomes available on either side of a breakpoint to pinpoint its location, resulting in shorter intervals.
For instance, the breakpoint intervals tend to be the smallest for events involving recombination
between the deeply divergent Omicron BA.1 and BA.2 sublineages (the red points in \ref{fig:recombinant_intervals}).

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020}
to relate the timing of recombination events with global case numbers.
Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in \ref{fig:recombinants_over_time}).
These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset.
Alongside the case numbers,
the black curve shows the number of recombinants inferred within each week along the ARG,
focusing on the 356 ``robust'' events.
Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high.
Overall, the number of recombinant events rose with the global number of cases
(Figure \ref{fig:recombinants_over_time}B; linear regression: $1.0 + 2.7 \; 10^{-7} x$, $p = 1.1 \; 10^{-7}$).
Recombination events also rose with the diversity of lineages in each week
(Figure \ref{fig:recombinants_over_time}C; linear regression: $0.9 + 4.8 x$, $p = 3.4\;10^{-7}$),
measured as the chance that two randomly drawn viruses have different Scorpio designations
($H$, referred to as the ``expected heterozygosity'').

We then asked which single measure best explains the number of recombination events inferred each week,
considering $n$, $H$, $n \; H$, $n^2$, $n^2 H$.
We included the squared number of cases each week thinking that it might better predict co-infections.
The best single predictor of recombination events involved the product of cases and expected heterozygosity ($n \; H$),
with an adjusted $R^2$ of 23.7\%.
This measure of case diversity ($n \; H$) explained 49\% more of the variation in recombination events per week
than the next best predictor ($H$, adjusted $R^2$ of 15.9\%).
While consistent with the hypotheses that recombination events occur more often when infection rates are common
and are detected more often when viral diversity is high, we note several sources of uncertainty.
Global case numbers were underreported, particularly during the Delta and Omicron waves \cite{Wang2025}.
Furthermore, there may be substantial uncertainty in the timing of a recombination event,
particularly if these occur in chronically infected individuals (as reported by \cite{Burel2022})
with substantial delays before onward cases are detected.
Finally, the filters that we applied removed all but the most strongly supported recombination events,
although similar conclusions were obtained when all 855 recombinants were considered (STAR Methods).

>>>>

\subsection*{Quantification of evidence for recombination}
%%% SHZ: We may not need this section at all, but I'm keeping it for now.
%%% We should probably have a condensed version of the first paragraph in the Pango X section.
%%% The second and third paragraphs should probably be removed entirely.
Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination, as measured by the number of resulting 
mutations (Figure~\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination, the stronger the 
evidence for recombination.
We calculated the number of averted mutations for all the ``robust'' recombination 
events (STAR Methods),
which include all the Pango X-associated recombination events
(Table~\ref{tab:pango_x_lineages}) and those studied by Jackson et
al.\cite{Jackson2021} (Table \ref{tab:jackson}).

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{fig_recombinant_evidence.pdf}
    \caption{Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes, colored according
to their classification by rebar (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labelled starting with `X' and
those identified by Jackson et al.\cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1). 
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
    \label{fig:averted_mutations}
\end{figure}

\subsection*{Recombination breakpoint intervals along the genome}
We explored the spatial distribution of the breakpoint locations of
the ``robust'' recombination events along the genome, and quantified how precisely breakpoints were localized.
For each event, we computed an interval estimate of its breakpoint location,
using its size as a measure of precision (median, 2,207 bases; STAR Methods).

We find an overall preponderance of breakpoint intervals towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene
(Figure~\ref{fig:recombinant_intervals}A,B),
consistent with patterns observed in previous 
studies\cite{Turakhia2022,Li2024}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
allowing recombination endpoints to be estimated more precisely.

Furthermore, we find a negative relationship between
breakpoint interval size and the amount of divergence between recombinant parents,
which is measured by time to their MRCA (Figure~\ref{fig:recombinant_intervals}).
When the two parental lineages have diverged for longer periods of time (the rightmost points),
more information becomes available on either side of a breakpoint to pinpoint its location, resulting in shorter intervals.
For instance, the breakpoint intervals tend to be the smallest for events involving recombination
between the deeply divergent Omicron BA.1 and BA.2 sublineages 
(the red points in Figure~\ref{fig:recombinant_intervals}).

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020}
to relate the timing of recombination events with global case numbers.
Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in 
Figure~\ref{fig:recombinants_over_time}).
These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset.
Alongside the case numbers,
the black curve shows the number of recombinants inferred within each week along the ARG,
focusing on the 356 ``robust'' events.
Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high.
Overall, the number of recombinant events rose with the global number of cases
(Figure~\ref{fig:recombinants_over_time}B; 
% JK is this correct?? we had 1.0 + 2.7 \; 10^{-7} x$ which I found confusing
linear regression: $1.0 + 2.7\times 10^{-7} x$, $p = 1.1 \times10^{-7}$).
Recombination events also rose with the diversity of lineages in each week
(Figure \ref{fig:recombinants_over_time}C; linear regression: 
$0.9 + 4.8 x$, $p = 3.4\times10^{-7}$),
measured as the chance that two randomly drawn viruses have different Scorpio designations
($H$, referred to as the ``expected heterozygosity'').

We then asked which single measure best explains the number of recombination events inferred each week,
considering $n$, $H$, $n H$, $n^2$, $n^2 H$.
We included the squared number of cases each week thinking that it might better predict co-infections.
The best single predictor of recombination events involved the product of cases and 
expected heterozygosity ($n H$),
with an adjusted $R^2$ of 23.7\%.
This measure of case diversity ($n H$) explained 49\% more of the variation in recombination events per week
than the next best predictor ($H$, adjusted $R^2$ of 15.9\%).
While consistent with the hypotheses that recombination events occur more often when infection rates are common
and are detected more often when viral diversity is high, we note several sources of uncertainty.
Global case numbers were underreported, particularly during the Delta and Omicron waves \cite{Wang2025}.
Furthermore, there may be substantial uncertainty in the timing of a recombination event,
particularly if these occur in chronically infected individuals (as reported by \cite{Burel2022})
with substantial delays before onward cases are detected.
Finally, the filters that we applied removed all but the most strongly supported recombination events,
although similar conclusions were obtained when all 855 recombinants were considered (STAR Methods).
